Density dependence

### 2020-09-15 R/competition wrangling.Rmd
add derived data for density calculations
changed density to be #shoots per cell

# Calculating density at different distances
First, a small example as proof of concept:
```{r}
ha_test <-
  ha %>%
  filter(plot == first(plot)) %>%
  filter(year == first(year)) %>%
  filter(row_number <=3, column <=3) %>% 
  add_column(plant = 1) #just a dummy indicator column for use later
ha_test
```

@@

```{r}
plant_summary <-
  ha_test %>% 
  group_by(row_number, column) %>% 
  summarize(n_plants = n()) %>% 
  rename(row = row_number)
plant_summary
```
Now I can join these two data frames.

@@

```{r}
#density between 5m and 15m away
dist_5_15<-
  dist_full %>% 
  filter(dist >=5, dist <15) %>% 
  group_by(row_from, column_from) %>% 
  summarize(n_neighbors_5_15 = sum(n_plants, na.rm = TRUE),
            n_cells = n(),
            density_5_15 = n_neighbors_5_15/n_cells)
```
So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2

@@

Now, create that summary tibble counting the number of plants in each grid cell for every plot:year combo, join, and unnest.

```{r}
ha_dist <-
  ha_nest_dist %>% 
  #summarize each year:plot to count number of plants in each grid cell
  mutate(n_plants_sum = map(data, ~{
    .x %>% 
      add_column(plant = 1) %>% 
      group_by(row_number, column) %>% 
      summarize(n_plants = n()) %>% 
      rename(row = row_number)
  }
  )) %>% 
  #join with tidied distance matrix to get dataframe for distance from every row
  #and column to every other row and column
  mutate(dist_full = map2(dist_df, n_plants_sum, ~{
    full_join(.x, .y, by = c("row_to" = "row", "column_to" = "column"))
  } 
  )) %>% 
  select(plot, year, dist_full) %>% 
  unnest(dist_full)

@@

```{r}
dens_0_10 <-
  ha_dist %>% 
  filter(dist>0 & dist <=10) %>% 
  group_by(plot, year, row_from, column_from) %>% 
  summarize(density_0_10 = sum(n_plants, na.rm = TRUE)/n())
left_join(ha, dens_0_10, by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
  arrange(plot, year, row_number, column)
```

@@

out <- 
  map2(L, U, ~{
  ha_dist %>%
    filter(dist >= .x & dist < .y) %>% 
    group_by(plot, year, row_from, column_from) %>%
    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_plants, na.rm = TRUE)/n())
})

@@

>>>>

# Calculating density at different distances
First, a small example as proof of concept:

```{r}
ha_test <-
  ha %>%
  filter(plot == first(plot)) %>%
  filter(year == first(year)) %>%
  filter(row_number <=3, column <=3) %>% 
  add_column(plant = 1) #just a dummy indicator column for use later
ha_test
```

@@

```{r}
plant_summary <-
  ha_test %>% 
  group_by(row_number, column) %>% 
  summarize(n_plants = n(),
            n_shts = sum(shts)) %>% 
  rename(row = row_number)
plant_summary
```
Now I can join these two data frames.

@@

```{r}
#density between 5m and 15m away
dist_5_15 <-
  dist_full %>% 
  filter(dist >=5, dist <15) %>% 
  group_by(row_from, column_from) %>% 
  summarize(n_neighbors_5_15 = sum(n_plants, na.rm = TRUE),
            n_cells = n(),
            density_5_15 = n_neighbors_5_15/n_cells,
            shtsper_5_15 = sum(n_shts, na.rm = TRUE)/n())
```
So at [1,1] there are only 3 cells between 5 and 15m away, there are a total of 6 plants in those cells, so the conspecific density at that distance is 2.0 plants / 100m^2 or 3.66 shoots / 100m^2

@@

Now, create that summary tibble counting the number of shoots in each grid cell for every plot:year combo, join, and unnest.

```{r}
ha_dist <-
  ha_nest_dist %>% 
  #summarize each year:plot to count number of plants in each grid cell
  mutate(n_shts_sum = map(data, ~{
    .x %>% 
      # add_column(plant = 1) %>% 
      group_by(row_number, column) %>% 
      summarize(n_shts = sum(shts)) %>% 
      rename(row = row_number)
  }
  )) %>% 
  #join with tidied distance matrix to get dataframe for distance from every row
  #and column to every other row and column
  mutate(dist_full = map2(dist_df, n_shts_sum, ~{
    full_join(.x, .y, by = c("row_to" = "row", "column_to" = "column"))
  } 
  )) %>% 
  select(plot, year, dist_full) %>% 
  unnest(dist_full)

@@

```{r}
dens_0_10 <-
  ha_dist %>% 
  filter(dist>0 & dist <=10) %>% 
  group_by(plot, year, row_from, column_from) %>% 
  summarize(density_0_10 = sum(n_shts, na.rm = TRUE)/n())
left_join(ha, dens_0_10, by = c("plot", "year", "row_number" = "row_from", "column" = "column_from")) %>% 
  arrange(plot, year, row_number, column)
```

@@

out <- 
  map2(L, U, ~{
  ha_dist %>%
    filter(dist >= .x & dist < .y) %>% 
    group_by(plot, year, row_from, column_from) %>%
    summarize(!!glue::glue("density_{.x}_{.y}") := sum(n_shts, na.rm = TRUE)/n())
})

@@

# Write to cleaned data

```{r}
write_rds(ha_new, here("analysis", "data", "derived_data", "ha_density.rds"))
```

### 2020-09-15 R/density dependence.Rmd
changed density to be #shoots per cell

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(lme4)
library(bbmle)
library(car)
library(broom)
library(broom.mixed)

@@

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```
Filter to just continuous forest
```{r}
ha_cf <- 
  ha %>% 
  filter(habitat == "CF") %>% 
  mutate(log_ht = log(ht),
         log_ht_prev = log(ht_prev))
```

I don't know what a height of 0 means...

```{r}
ha_cf <- ha_cf %>% filter(ht > 0, ht_prev > 0)
```

@@

```{r}
m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
summary(m)
m2 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year), data = ha_cf)
m3 <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|plot), data = ha_cf)
m4 <- lm(log_ht ~ log_ht_prev + density_0_10, data = ha_cf)
AICtab(m, m2, m3, m4)
```

keep both random effects, possibly remove random effect of plot.

```{r}
m <- lmer(log_ht ~ log_ht_prev + density_0_10 + (1|year) + (1|plot), data = ha_cf)
Anova(m)
```

Density matters?

```{r}
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()
```

residuals look leptokurtic.

```{r}
augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```

Try scaled T distribution

```{r}
library(mgcv)
library(gamm4)
m <- gamm4(log_ht ~ log_ht_prev + density_0_10, random = ~ (1|year) + (1|plot), family = scat, data = ha_cf)
m <- gam(log_ht ~ log_ht_prev + density_0_10, family = scat, data = ha_cf)
coef(m)
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()

augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```

Maybe something like a gamma distribution on un-transformed values?  Or use shoots..

```{r}
m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + (1|year) + (1|plot), data = ha_cf)
Anova(m)
augment(m) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()

augment(m) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()
```
Ok, yeah.  That's a good reason to use # of shoots maybe?

```{r}
Anova(m)
```

Density matters!

```{r}
m <- lmer(log(shts) ~ log(shts_prev) * density_0_10 + (1|year) + (1|plot), data = ha_cf)
fixef(m)
Anova(m)
```

Interaction doesn't! (bless)

```{r}
m <- lmer(log(shts) ~ log(shts_prev) + density_0_10 + density_10_20 + density_20_30 + (1|year) + (1|plot), data = ha_cf)
fixef(m)
Anova(m)
```

Ah, density further away matters as well.  Collinearity, probably, so don't trust p-values!

Ok, gotta do GAMs I guess!

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(lme4)
library(bbmle)
library(car)
library(broom)
library(broom.mixed)
library(patchwork)

@@

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```

Filter to just continuous forest

```{r}
ha_cf <- 
  ha %>% 
  filter(habitat == "CF") %>% 
  mutate(log_ht = log(ht),
         log_ht_prev = log(ht_prev))
```

I don't know what a height of 0 means...
Height of 0 probably means it was knocked over, or very small, or dead.

Scale year

```{r}
ha_cf <- 
  ha_cf %>% 
  ungroup() %>% 
  mutate(year_post = year - min(year))
```

@@

For this I'm going to use shoot number. Very closely related to height, but better behaved distribution and not sensitive to plants being knocked over.

```{r}
mpois <- glm(shts ~ log(shts_prev) * density_0_10,
             family = poisson(link = "log"),
             data = ha_cf)
lm <- lm(shts ~ log(shts_prev) * density_0_10,
         data = ha_cf)
mnorm <- glm(shts ~ log(shts_prev) * density_0_10,
             family = gaussian(link = "log"),
             data = ha_cf)
mpois2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
                family = poisson(link = "log"),
                data = ha_cf)

mnorm2 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post) + (1|plot),
                family = gaussian(link = "log"),
                data = ha_cf)

AICtab(mpois, mnorm, lm, mpois2, mnorm2)
```

Lognormal distribution fits best, even after attempting to account of overdispersion with random effects.  I guess despite being discrete, this isn't really "count" data *per se*.

Model doesn't converge though.

```{r}
mnorm3 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number) + (1|year_post),
               family = gaussian(link = "log"),
               data = ha_cf)

AICtab(mnorm2, mnorm3)
```

Can remove plot ID.

```{r}
mnorm4 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|year_post),
               family = gaussian(link = "log"),
               data = ha_cf)
AICtab(mnorm3, mnorm4)
```

Keep plant ID as a random effect.

```{r}
mnorm5 <- glmer(shts ~ log(shts_prev) * density_0_10 + (1|ha_id_number),
               family = gaussian(link = "log"),
               data = ha_cf)
AICtab(mnorm3, mnorm4, mnorm5)
```

Can't remove year as random effect.

Inspect residuals:

```{r}
h <- augment(mnorm2) %>% 
  ggplot(aes(.resid)) +
  geom_histogram()

qq <- augment(mnorm2) %>% 
  ggplot(aes(sample = .resid)) +
  geom_qq() + geom_qq_line()

resid<- augment(mnorm2) %>% 
  ggplot(aes(x = .fitted, y = .resid)) + geom_point() 
h/(qq|resid)
```
I don't love this, but it is a better fit than a poisson distribution.

# TODO:

- Density should be total shoot number
- Should actually be density in previous year
- ~~Shoots should be poisson distribution, not log-transformed?~~
- incorporate 1-ha fragments and use habitat type as a predictor.
- Implement FLM approach to density dependence.
